Co-evolution of lexical meaning \& pragmatic use

### 2017-03-23 paper/paper-v3.tex
revisions up to general discussion

Expressivity and learnability are central to the cultural evolution of language. These
components can be modelled, respectively, as replication based on a measure of fitness in terms
of communicative efficiency and iterated Bayesian learning. Their interaction is described by
the discrete time replicator mutator dynamic in (\ref{eq:RMD_discrete}), repeated here:
\begin{align*}
  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_k x_k f_k}\,.
\end{align*}
This equation defines the frequency $x'_i$ of type $i$ at the next time step, based on its
frequency $x_i$ before the step, its fitness $f_i$, and the probability that a learner infers
$i$ when observing the behavior of a type-$j$ agent. Fitness-based selection can be thought of
as biological (fitness as expected relative number of offspring) or cultural (fitness as 
likelihood of being imitated or repeated). The types that the dynamic operates on are pairs
consisting of a lexicon and a pragmatic use pattern. A type's expressivity depends on its
communicative efficiency within a population while its learnability depends on the fidelity by
which it is inferred by new generations of learners. The learners' task is consequently to
perform a joint inference over types of linguistic behavior and lexical meaning.

\section{Scalar implicatures}
\label{sec:si-case-study}

Scalar implicatures are a particularly well-studied type of conventional pragmatic
inferences. They are licensed for groups of expressions ordered in terms of informativity, here
understood as an entailment induced order. For instance, {\em some} is entailed by {\em
  all}. If it were true that ``Chris owns all of Johnny Cash's albums'', it would also be true
that ``Chris owns some of Johnny Cash's albums''. However, while weaker expressions such as
{\em some} are truth-conditionally compatible with stronger alternatives such as {\em all},
this is not what their use is normally taken to convey. Instead, the use of a less informative
expression when a more informative one could have been used can license a defeasible inference
that stronger alternatives do not hold (cf. \citealt{horn:1972,gazdar:1979}). That is, a hearer
who assumes the speaker to be able and willing to provide all relevant information can infer
that stronger alternatives do not hold because the speaker used a weaker alternative
instead. In this way, ``Chris owns some of Johnny Cash's albums'' is strengthened to convey
that he owns {\em some but not all} albums. A bound that rules out stronger alternatives is
thusly not codified in the lexical meaning of weak alternatives but instead pragmatically
supplied.

As noted earlier, this kind of strengthening is captured by the linguistic behavior of pragmatic types introduced in \S\ref{sec:languages+use}: A pragmatic hearer who reasons about the use of a message involving a weak scalar alternative will associate it more with a state in which stronger alternatives do not hold. This is so because a rational speaker would use a more informative message when in such a state. Conversely, a pragmatic speaker will reason about her interlocutor's expected interpretation and use the messages at her disposition accordingly as to maximize communicative success.

Our initial question about the division of labor between semantics and pragmatics can be narrowed to the case of scalar implicatures by asking for a justification for the lack of lexical upper-bounds in weak scalar alternatives. That is, we ask why lexical meanings that lack upper-bounds and convey it pragmatically are regularly selected for over alternatives such as that of codifying the bound semantically. More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} and {\em big} were truth-conditionally incompatible with stronger alternatives such as {\em hot}, {\em and}, {\em all} and {\em huge} to avoid misunderstanding?  This question is particularly striking considering the number of expressions that license such inferences across natural languages.

\subsection{Analysis}
The dynamics are initialized with an arbitrary distribution over types, constituting a population's first generation. In the following we report the outcome of their development under different parameters and pressures after $50$ generations. This ensures that these outcomes correspond to developmental plateaus after which no noteworthy change is registered. As specified in \S\ref{sec:learnability}, the mutation matrix $Q$ can be obtained by considering all possible state-message sequences of length $k$. Given that this is intractable for large $k$, the sets of data learners are exposed to are approximated by sampling $250$ $k$-length sequences from each type's production probabilities.

We begin setting the stage by defining the learning prior and the set of types under consideration. Drawing from our preceding discussion, functional pressure on successful communication combined with learning pressures in the form of a bias that favors simpler semantics may lead to the selection of $\Llack$-like semantics and to a division of labor between semantics and pragmatics. We first showcase the effects that expressivity and learnability have in isolation before turning to those resulting from their combination.

\subsubsection{An inductive learning bias for semantic simplicity}
There is a growing effort to develop empirically  testable representational  languages that allow for the measure of semantic complexity. For instance, so-called {\em languages of thought} (LOT) have been put to test in various rational probabilistic models that show encouraging results (see, e.g., \citealt{katz+etal:2008, piantadosi+etal:underreview, piantadosi+etal:2012a} and \citealt{piantadosi+jacobs:2016} for recent discussion). At its core, a LOT defines a set of operations and composition rules from which lexical meaning can be derived. As a first approximation and for the sake of concreteness, we follow this approach to make the preference of learners for simpler semantic representations precise and define the complexity of a representation as a function of its derivation cost in a weighed generative LOT.

Our toy grammar of concepts is given in Table~\ref{tab:grammar}. This grammar uses basic set-theoretic operations to form expressions which can be evaluated as true or false in different world states such as $\ssome$ or $\sall$. Applications of generative  rules have a cost attached to them. Here we simply assume that Boolean combinations of concepts are more complex than atomic concepts and that otherwise each rule application adds the same cost unit.
\begin{table}
  \centering
  \begin{align*}
    C & \rightarrow_2 C \wedge C 
    & 
    X & \rightarrow_1 \set{A,B} \\
    C & \rightarrow_2 \neg C 
    & 
    X & \rightarrow_1 X \cap X \\
    C & \rightarrow_1 X \subseteq X
    & 
    X & \rightarrow_1 X \cup X \\
    C & \rightarrow_1 X \neq \emptyset \\
    C & \rightarrow_1 X = \emptyset     
  \end{align*}
  \caption{Toy grammar in a set-theoretic LOT with weighted rules.}
  \label{tab:grammar}
\end{table}

The prior probability of a type is just the prior probability of its lexicon. The prior of a lexicon is a function of the complexity of the lexical representations in its image set. As motivated above, lexica that use simpler concepts are \emph{a priori} more likely. A simple way of defining such priors over lexica is:
\begin{align*}
  P(L)  & \propto \prod_{c \in Im(L)} P(c)   \ \text{, with} & 
  P(c) & \propto \max_{c'}Compl(c') - Compl(c) + 1\,,
\end{align*}

where $Compl(c)$ is the complexity of the minimal derivation length of $c$ according to the LOT-grammar. There are many ways to define priors over lexica (see, e.g., \citealt{goodman+etal:2008, piantadosi+etal:2012a}) but the key assumption here, common to all of them, is that simple representational expressions should be favored over more complex ones. We should stress that these details -- from the generative grammar to its complexity measure -- are to be regarded as an instantiation of our general assumptions instead of as empirical claims in their own right.

\subsubsection{Lexica, signaling behavior \& types}
We consider a state space with three states $\States = \set{\snone, \ssome, \sall}$. \tb{Should we say something about the inclusion of $\snone$? It may be more natural to introduce this state earlier but may unnecessarily complicate the exposition of signaling behavior in the previous section.} This space can be
thought of as a partition of possible worlds into cells where none, some or all of the $A$s are
$B$s, for some arbitrary fixed predicates $A$ and $B$. Eight concepts can be distinguished
based on their truth or falsity in three world states, six of which are not contradictory
 or tautological. These are listed with mnemonic names in
Table~\ref{tab:concepts} together with their complexity according to the grammar in Table \ref{tab:grammar}.

@@

A lexicon $L$ is a mapping $\Messgs \rightarrow C$ from messages to concepts. With three
messages there are $6^3 = 216$ possible lexica. Some assign the same concept to more than one message and others lexicalize the same concepts but associate them with different messages. The main results reported in the following do not hinge on the inclusion of these lexica but this unaltered space does better illustrate the influence and differences of the linguistic pressures we consider. Additionally, the competition of different languages that lexicalize the same concepts through different forms allows us to take serious the problem of only one being selected for even though multiple alternatives of equal expressivity {\em and} learnability exist.

Out of these possible lexica, three kinds are of particular relevance. First, lexica that assign the same concept to more than one message. Such lexica lack in expressivity but may be favored by the learning prior should they lexicalize simple concepts. Second, lexica that conventionalize upper-bounds to realize a (quasi-)partition of the state space. As discussed before in relation to $\Lbound$, users of these lexica need not resort to pragmatic reasoning to convey an upper-bound with weak scalar expressions. Instead, this information is already provided by the semantics of their lexicon, giving them an advantage in terms of expressivity. Types with such lexica are expected to be the main contenders of the natural language like semantics that do not lexicalize an upper-bound for weak scalar expressions. Third and lastly, $\Llack$-style lexica that, paired with mutual reasoning, can convey an upper-bound pragmatically. There are six lexica of the second kind and six of the third. The following three lexica exemplify each kind, with $\Lall$ being a lexicon of the first kind. This lexicon is not optimal for communication as it assigns all its messages to the same state, $\sall$. This inevitably leads to a communicative disadvantage in its use. However, it may have a learnability advantage over our target $\Llack$ and its competitor $\Lbound$ in virtue of being the simplest lexicon, making it a priori favored by learners.

\begin{align*}
  \begin{blockarray}{lccc}
    & & \underline{\Lall} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
     \mygray{\snone}  & 0 & 0 & 0\\
     \mygray{\ssome}  & 0 & 0 & 0\\
    \mygray{\sall}   & 1 & 1 & 1\\
    \end{block}
% \\ & & P(\Lall)\rlap{ $\propto 9$} & \\
  \end{blockarray} & &
 \begin{blockarray}{lccc}
    & & \underline{\Lbound} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
       & 1 & 0 & 0\\
       & 0 & 1 & 0\\
       & 0 & 0 & 1\\
    \end{block}
%  \\& & P(\Lbound)\rlap{ $\propto 96$} & \\
  \end{blockarray} & &
  \begin{blockarray}{lccc}
    & & \underline{\Llack} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
       & 1 & 0 & 0\\
       & 0 & 1 & 0\\
       & 0 & 1 & 1\\
    \end{block}
%   \\& & P(\Llack)\rlap{ $\propto 48$} & \\
  \end{blockarray}
\end{align*}

Recall that types are a combination of a lexicon and a manner of language use. We analyze the model's predictions in populations of types with one of the two behaviors introduced earlier; literal or pragmatic. The former correspond to level $0$ reasoners and the latter to ones of level $1$. Higher level reasoning is not required to derive scalar implicatures from these lexica, nor do they leave room for substantial pragmatic refinement. Accordingly, we consider a total of $432$ types. Six are variants of our pragmatic $\Llack$ target and twelve are either literal or pragmatic speakers of variants of its main contender, $\Lbound$.

%Combining a linguistic behavior with each of these $6$ lexica yields a total of $12$ distinct types. , such as $L_4$, will produce speaker behavior that is {\em almost} indistinguishable from that of a language that lacks upper-bounds, but with pragmatic users, such as $L_5$. Almost, because there may be slight differences between the probability with which speakers would (erroneously) use a semantically false description and the probability with which speakers would (erroneously) use a pragmatically suboptimal description. Due to this possibly marginal difference between pragmatic $L_4$ and $L_5$, the selection of one type over the other is expected to mainly depend on how well each can be transmitted to new learners. Things are less clear for literal $L_5$ contrasted with literal/pragmatic $L_4$. The former has a learning advantage when learners are biased against upper-bounds but is expected to fare worse in communicative terms.

\subsubsection{Expressivity only} 
\begin{figure}
\centering
\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/fig1-onlyr}
\caption{Stacked proportion of pragmatic $\Llack$-style types and incumbent types, when other than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under only a pressure for expressivity. The right-most plots zoom in on only the proportion of pragmatic $\Llack$ types.}
\label{fig:only-R}
\end{figure}

Expressivity is sensitive to $\lambda$ as it influences signaling behavior. This is showcased in Figure \ref{fig:only-R}, which depicts the proportion of pragmatic $\Llack$-style speakers in $5$ independent populations after $50$ generations, as well as that of the type with the highest proportion in a given population, the incumbent. Low $\lambda$ levels communicative differences between types, as even those that can differentiate between all states semantically or pragmatically do not always maximize expected communicative success. The outcome is that of a mixed population that allows for the survival of competing types. Conversely, higher $\lambda$ promotes more rational linguistic behavior, widening the gap in expressivity between types and promoting more homogeneous populations. As suggested by Figure \ref{fig:only-R}, the incumbent in most populations is not one of the six pragmatic $\Llack$-style types. That is, a pressure only for expressivity does not lend a justification to the systematic prevalence of a lack of upper-bounds under any $\lambda$-value. For instance, in $1000$ independent populations with $\lambda = 20$ pragmatic $\Llack$ types only made up $11$ cases of incumbency, corresponding to a mean proportion of $0.003$ across populations. By contrast, $913$ incumbents were $\Lbound$-style users with close to an even share between literal ($454$) and pragmatic types ($459$), corresponding to a mean proportion of about $.48$ taken together. %Across types and populations incumbents make up $.50$ of the mean outcome.

\subsubsection{Learnability only}

@@

The effect of iterated learning without a pressure for expressivity using either posterior sampling ($l = 1$) or a stronger tendency towards posterior maximization ($l = 15$) is shown in Figure \ref{fig:only-M} together with the prior over types. The prior shows that while users of $\Llack$ are not the most favored by the inductive bias (compared, e.g., to $\Lall$) they are nevertheless more advantaged than others, such as $\Lbound$, in virtue of the relatively simple semantics they conventionalize. Crucially, $\Llack$ enables its users to convey each state with a single message when combined with pragmatic reasoning provided sufficiently high $\lambda$. This makes it less likely to be confused with other types if the learning data is not too sparse ($k \geq 5$). Put differently, learners have a higher propensity to infer pragmatic $\Llack$ over types that might be hard or impossible to distinguish it from based on overt linguistic behavior, such as $\Lbound$, and $\Llack$ is also less likely to be mistaken with types with different observable behavior because its pragmatic use approximates a one-to-one form-meaning mapping. As a consequence, a stronger propensity to maximize the posterior increases their proportion in the population.

However, in contrast to a pressure only for expressivity with high $\lambda$, learnability alone does not succeed in selecting for a single type of a kind. For instance, for only one of the six pragmatic $\Llack$ types. Instead, it leads to their coexistence. Each is passed on to next generation with the same faithfulness and, differently from a pressure for communicative success, they do not stand in competition with each other. As implicit in Figure \ref{fig:only-M}, in $1000$ independent populations all incumbents were pragmatic $\Llack$ users provided sufficiently high $\lambda$, with each reaching approximately the same proportion of users in the population. As with expressivity only, low values of $\lambda$ make the differences in observable behavior across types less pronounced and therefore reflect the learners' inductive bias more faithfully, favoring functionally deficient but a priori preferred types such as those that use $\Lall$. A pressure for learnability alone may consequently lead to a spread of communicatively suboptimal types that are easier to learn. In the extreme, when $l = 1$ and $\lambda = 1$ all of $1000$ independent populations had users of $\Lall$ as incumbents.

\subsubsection{Expressivity and learnability}
The separate inspection of each pressure not only showcases the contribution of the model's components but also highlights some of their broader implications for the emergence of a division of labor between semantics and pragmatics. Neither dynamic on its own comes close to converging to a monomorphic population of pragmatic $\Llack$ speakers. When pressured toward expressivity, the slight communicative advantage of $\Lbound$'s lexicalization of an upper-bound over pragmatic use of $\Llack$ leads to the former's selection. When pressured toward learnability, pragmatic $\Llack$ is promoted over functionally similar but semantically more complex alternatives such as $\Lbound$. However, on its own learnability does not foment the propagation of these types across the population. What is more, it leads to heterogeneous populations that foster an equal share of types of a kind, such as different pragmatic $\Llack$ types.

@@

To recapitulate, other than the involvement of both pressures, the resulting proportion of pragmatic $\Llack$ speakers primarily hinges on three factors. First, the degree to which agents try to maximize communicative success. This plays a role both in expressivity as well as in producing data that allows learners to discriminate this type from others. Second, the inductive bias, which leads learners to prefer simpler over more complex semantic representations in acquisition. Lastly, the posterior parameter, as it magnifies the effects of the learning bias in tandem with replication.

\subsection{Discussion}
Under the assumption of a learning bias for simpler lexical representations, our results suggest that a lack of semantic upper-bounds coupled with pragmatic reasoning can overcome communicative pressures and be majoritarily selected for in a population. This prediction hinges on three assumptions. First, that language is pressured toward both expressivity and learnability. Second, that language users favor expressions/interpretations expected to have a higher likelihood of being understood/meant over ones with a lower likelihood. Lastly, that learners prefer simpler over more complex lexical representations and exhibit a tendency towards the acquisition of the type that best explains the learning data. This outcome is particularly encouraging in light of other advantages a lack of semantic upper-bounds may confer, as discussed at the beginning of this section.

>>>>

Expressivity and learnability are central to the cultural evolution of language. These
components can be modelled, respectively, as replication based on a measure of fitness in terms
of communicative efficiency and iterated Bayesian learning. Their interaction is described by
the discrete time replicator mutator dynamic in (\ref{eq:RMD_discrete}), repeated here:
\begin{align*}
  x'_i = \sum_j Q_{ji} \frac{x_jf_j}{\sum_k x_k f_k}\,.
\end{align*}
This equation defines the frequency $x'_i$ of type $i$ at the next time step, based on its
frequency $x_i$ before the step, its fitness $f_i$, and the probability that a learner infers
$i$ when observing the behavior of a type-$j$ agent. Fitness-based selection is here thought of
not as biological (fitness as expected relative number of offspring) but cultural (fitness as
likelihood of being imitated or repeated) evolution, since the types that the dynamic operates
on are pairs consisting of a lexicon and a pragmatic use pattern. A type's expressivity depends
on its communicative efficiency within a population while its learnability depends on the
fidelity by which it is inferred by new generations of learners. The learners' task is
consequently to perform a joint inference over types of linguistic behavior and lexical
meaning.

The model has three parameters: $\lambda$ regulates the degree to which pragmatic speakers
choose messages that maximize the chance of communicative success; $k$ is the number of
observations for each language learner; $l$ regulates where the learners' inference behavior
lies on a spectrum from probability matching to choice of the most likely parent type.

\section{Case study: scalar implicatures}
\label{sec:si-case-study}

The model of the previous section formalizes the evolutionary competition between different
pairs of lexical concepts and ways of using them. This section looks at a concrete case study
on scalar implicatures. Concretely, we will engage in the following formal thought experiment:
if a population of language users could freely combine different lexica with different
pragmatic strategies, what there the conditions under which the majority view of scalar
implicatures could have evolved?

Recall that the majority view is that scalar implicatures are non-lexicalized pragmatic
enrichments. Scalar implicature triggers like \emph{some}, \emph{warm} or \emph{may} are
semantically weak expressions for which logically stronger expressions are salient, e.g.,
\emph{all}, \emph{hot} or \emph{must}. For instance, {\em some} is entailed by {\em all}. If
the sentence ``Chris owns all of Johnny Cash's albums'' is true, then ``Chris owns some of
Johnny Cash's albums'' is also true. However, while weaker expressions such as {\em some} are
truth-conditionally compatible with stronger alternatives such as {\em all}, this is not what
their use is normally taken to convey. Instead, the use of a less informative expression when a
more informative one could have been used can license a defeasible inference that stronger
alternatives do not hold (cf. \citealt{horn:1972,gazdar:1979}). That is, a hearer who assumes
the speaker to be able and willing to provide all relevant information can infer that stronger
alternatives do not hold because the speaker used a weaker alternative instead. In this way,
``Chris owns some of Johnny Cash's albums'' is strengthened to convey that she owns {\em some
  but not all} albums. The upper bound, so the majority view, is a pragmatic inference but not
part of the conventional meaning. \mf{I wonder if this paragraph is really still necessary
  after the introduction?}

In the following we consider a specific application of the model from Section~\ref{sec:model}
which allows to address the question if or when scalar inferences might (not) lexicalize. We
consider what is perhaps one of the simplest non-trivial setups that speak to this matter and
reflect on its limitations in Section~\ref{sec:discussion}. The setup is introduced in
Section~\ref{sec:setup:-stat-conc}. Section~\ref{sec:simulation-results} describes the simulations
and their results.

% Our initial question about the division of labor between semantics and pragmatics can be narrowed to the case of scalar implicatures by asking for a justification for the lack of lexical upper-bounds in weak scalar alternatives. That is, we ask why lexical meanings that lack upper-bounds and convey it pragmatically are regularly selected for over alternatives such as that of codifying the bound semantically. More poignantly, would it not serve language users better if weak(er) expressions such as {\em warm}, {\em or}, {\em some} and {\em big} were truth-conditionally incompatible with stronger alternatives such as {\em hot}, {\em and}, {\em all} and {\em huge} to avoid misunderstanding?  This question is particularly striking considering the number of expressions that license such inferences across natural languages.

\subsection{Setup}
\label{sec:setup:-stat-conc}

To fill the model from Section~\ref{sec:model} with life, we need to specify the sets of
states, messages and lexica. Additionally, we want to explore the effects of a learning bias in
favor of simple lexical concepts. One way of motivating and formalizing such a bias is
introduced below.

\subsubsection{States, messages, concepts and Lexica}

Consider a state space with three states $\States = \set{\snone, \ssome, \sall}$ and think of
it as a partition of possible worlds into cells where none, some or all of the $A$s are $B$s,
for some arbitrary fixed predicates $A$ and $B$. Eight concepts can be distinguished based on
their truth or falsity in three world states, six of which are not contradictory or
tautological.

A lexicon $L$ is a mapping $\Messgs \rightarrow C$ from messages to concepts. With three
messages there are $6^3 = 216$ possible lexica. Some assign the same concept to more than one
message and others lexicalize the same concepts but associate them with different messages. Out
of these possible lexica, three kinds are of particular relevance. First, lexica that assign
the same concept to more than one message. Such lexica lack in expressivity but may be favored
by particular learning biases nonetheless. Second, lexica that conventionalize upper-bounds to
realize a (quasi-)partition of the state space. Finally, lexically that do not lexicalize an
upper bound but allow it to be conveyed pragmatically due to the presence of a stronger lexical
item. There are six lexica of the second kind and six of the third. The following three lexica
exemplify each kind:
\begin{align*}
  \begin{blockarray}{lccc}
    & & \underline{\Lall} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
     \mygray{\snone}  & 0 & 0 & 0\\
     \mygray{\ssome}  & 0 & 0 & 0\\
    \mygray{\sall}   & 1 & 1 & 1\\
    \end{block}
% \\ & & P(\Lall)\rlap{ $\propto 9$} & \\
  \end{blockarray} & &
 \begin{blockarray}{lccc}
    & & \underline{\Lbound} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
       & 1 & 0 & 0\\
       & 0 & 1 & 0\\
       & 0 & 0 & 1\\
    \end{block}
%  \\& & P(\Lbound)\rlap{ $\propto 96$} & \\
  \end{blockarray} & &
  \begin{blockarray}{lccc}
    & & \underline{\Llack} & \\
    & \mygray{\mnone} & \mygray{\msome} & \mygray{\mall} \\
    \begin{block}{l[ccc]}
       & 1 & 0 & 0\\
       & 0 & 1 & 0\\
       & 0 & 1 & 1\\
    \end{block}
%   \\& & P(\Llack)\rlap{ $\propto 48$} & \\
  \end{blockarray}
\end{align*}
Lexicon $\Lall$ is clearly bad for communication: all message and interpretation choices will
be equally likely for all types; no information will be conveyed with it. Users of $\Lbound$
need not resort to pragmatic reasoning to convey an upper-bound with weak scalar
expressions. Instead, this information is already provided by the semantics of their lexicon,
giving them an advantage in terms of expressivity. Types with such lexica may be expected to
generally achieve higher communicative success than pragmatic users with a lexicon of type
$\Llack$ (see Section~\ref{sec:languages+use}).

Recall that types are a combination of a lexicon and a manner of language use. We analyze the
model's predictions in populations of types with one of the two behaviors introduced earlier;
literal or pragmatic. The former correspond to level $0$ reasoners and the latter to ones of
level $1$. Accordingly, we consider a total of $432$ types. Six are variants of pragmatic
language users with $\Llack$-like lexica. These will be our target types, because they
represent lexica and language use that conform to the majority view of scalar
implicatures. Twelve types are either literal or pragmatic speakers with lexical of the
$\Lbound$ kind. These will be the target types main competitors in evolutionary competition.

\subsubsection{An inductive learning bias for semantic simplicity}
\label{sec:an-induct-learn}

There is a growing effort to develop empirically testable representational languages that allow
for the measure of semantic complexity. For instance, so-called {\em languages of thought}
(LOT) have been put to test in various rational probabilistic models that show encouraging
results (see, e.g., \citealt{katz+etal:2008, piantadosi+etal:underreview,
  piantadosi+etal:2012a} and \citealt{piantadosi+jacobs:2016} for recent discussion). At its
core, a LOT defines a set of operations and composition rules from which concepts can be
derived. As a first approximation and for the sake of concreteness, we follow this approach to
make the preference of learners for simpler semantic representations precise and define the
complexity of a representation as a function of its derivation cost in a weighed generative
LOT.

\begin{table}
  \centering
  \begin{align*}
    C & \rightarrow_2 C \wedge C 
    & 
    X & \rightarrow_1 \set{A,B} \\
    C & \rightarrow_2 \neg C 
    & 
    X & \rightarrow_1 X \cap X \\
    C & \rightarrow_1 X \subseteq X
    & 
    X & \rightarrow_1 X \cup X \\
    C & \rightarrow_1 X \neq \emptyset \\
    C & \rightarrow_1 X = \emptyset     
  \end{align*}
  \caption{Toy grammar in a set-theoretic LOT with weighted rules.}
  \label{tab:grammar}
\end{table}

Our toy grammar of concepts is given in Table~\ref{tab:grammar}. This grammar uses basic
set-theoretic operations to form expressions which can be evaluated as true or false in states
$\snone$, $\ssome$ or $\sall$ from above. Applications of generative rules have a cost attached
to them. Here we simply assume that Boolean combinations of concepts are more complex than
atomic concepts and that otherwise each rule application adds the same cost
unit. Table~\ref{tab:concepts} lists all six concepts relevant here, their truth conditions and
the simplest formula that expresses this concept from the grammar in Table~\ref{tab:grammar}.

@@

The complexity measure for lexical concepts from Table~\ref{tab:concepts} is used to define a
learning bias that favor simpler concepts over more complex ones. The prior probability of a
type is just the prior probability of its lexicon. The prior of a lexicon is a function of the
complexity of the lexical representations in its image set. Lexica with simpler concepts have a
higher prior. One simple way of defining such priors over lexica (and thereby types) is:
\begin{align*}
  P(L)  & \propto \prod_{c \in Im(L)} P(c)   \ \text{, with} & 
  P(c) & \propto \max_{c'}Compl(c') - Compl(c) + 1\,,
\end{align*}
where $Compl(c)$ is the complexity of the minimal derivation length of $c$ according to the
LOT-grammar (see Table~\ref{tab:concepts}). There are many ways to define priors over lexica
(see, e.g., \citealt{goodman+etal:2008, piantadosi+etal:2012a}) but the key assumption here,
common to all of them, is that simple representational expressions should be favored over more
complex ones. We should stress that these details -- from the generative grammar to its
complexity measure -- are to be regarded as one convenient operationalization of one general
approach to explicating learning biases; this is not a commitment that this general approach is
necessarily superior or that, within it, this particular instrumentalization is the single most
plausible.

\subsection{Simulation results}
\label{sec:simulation-results}

Recall that there are three parameters: soft-max parameter $\lambda$ affects how strongly
pragmatic speakers favor more informative messages; the bottleneck size $k$ influences how
faithfully learners can identify their teacher type; $l$ defines the learners disposition
towards choosing the most likely teacher type from the posterior distribution. We expect that
competitor types (types with lexica of the kind $\Lbound$) have a fitness advantage over target
types (pragmatic agents with lexica of the kind $\Llack$), especially for very low levels of
$\lambda$. Selection based on fitness alone may therefore not lead to prevalence of target
types in the population. On the other hand, lexica of type $\Llack$ are simpler than those of
type $\Lbound$ by the postulated measure from above. This may make them more likely to be
adopted by learners, especially when $k$ is low so that teacher types are relatively
indistinguishable based on their behavior in general and when $l$ is high. Still, lexica of the
kind $\Lall$ are in turn more likely \emph{a priori} than lexica of the kind
$\Llack$. Simulation results will shed light on the question whether target types can emerge
and for which parameter constellations.

To better understand the workings a pressure towards communicative efficiency and a pressure
towards learnability, we look at the behavior of the replicator and mutator step first in
isolation, and then in combination. All simulation runs are initialized with an arbitrary
distribution over types, constituting a population's first generation. All reported results are
the outcome of 50 update steps for each different constellation of parameter values. This
ensures that these outcomes correspond to developmental plateaus in which change is, if not
absent, then at least very slow, In other words, even if the resulting states do not correspond to
an eventual attracting state, they characterize long almost stationary states in which the
system remains for a very long time. As specified in \S\ref{sec:learnability}, the mutation
matrix $Q$ can be obtained by considering all possible state-message sequences of length
$k$. Given that this is intractable for large $k$, the sets of data learners are exposed to are
approximated by sampling $250$ $k$-length sequences from each type's production probabilities.

\subsubsection{Replication only: selection based on expressivity and communicative success} 
\begin{figure}
\centering
\includegraphics[width=\textwidth,height=8cm, keepaspectratio]{./plots/fig1-onlyr}
\caption{Stacked proportion of pragmatic $\Llack$-style types and incumbent types, when other than pragmatic $\Llack$, in $5$ independent populations after $50$ generations under only a pressure for expressivity. The right-most plots zoom in on only the proportion of pragmatic $\Llack$ types.}
\label{fig:only-R}
\end{figure}

As expected, selection based on communicative success is sensitive to $\lambda$ as it
influences signaling behavior. This is showcased in Figure \ref{fig:only-R}, which depicts the
proportion of target types in $5$ independent populations after $50$ generations, as well as
that of the type with the highest proportion in a given population, the incumbent. With low
$\lambda$ many types have very similar behavior, so that evolutionary selection lacks grip and
becomes very slow. The result is a very long transition with near stagnancy in a rather
homogeneous population with many types. Conversely, higher $\lambda$ promotes more rational
linguistic behavior, widening the gap in expressivity between types and promoting more
homogeneous populations. As suggested by Figure \ref{fig:only-R}, the incumbent in most
populations is not one of the six pragmatic $\Llack$-style types. That is, a pressure only for
expressivity does not lend a justification to the systematic prevalence of a lack of
upper-bounds under any $\lambda$-value. For instance, in $1000$ independent populations with
$\lambda = 20$ pragmatic $\Llack$ types only made up $11$ cases of incumbency, corresponding to
a mean proportion of $0.003$ across populations. By contrast, $913$ incumbents were
$\Lbound$-style users with close to an even share between literal ($454$) and pragmatic types
($459$), corresponding to a mean proportion of about $.48$ taken
together. %Across types and populations incumbents make up $.50$ of the mean outcome.

\subsubsection{Iterated learning only}

@@

The effect of iterated learning without a pressure for expressivity using either posterior
sampling ($l = 1$) or a stronger tendency towards posterior maximization ($l = 15$) is shown in
Figure \ref{fig:only-M} together with the prior over types. The prior shows that while users of
$\Llack$ are not the most favored by the inductive bias (compared, e.g., to $\Lall$) they are
nevertheless more advantaged than others, such as $\Lbound$, in virtue of the relatively simple
semantics they conventionalize. Crucially, $\Llack$ enables its users to convey each state with
a single message when combined with pragmatic reasoning provided sufficiently high
$\lambda$. This makes it less likely to be confused with other types if the learning data is
not too sparse ($k \geq 5$). Put differently, learners have a higher propensity to infer
pragmatic $\Llack$ over types that might be hard or impossible to distinguish it based on overt
linguistic behavior, such as $\Lbound$. Moreover, $\Llack$ is also less likely to be confused
with types with different observable behavior because its pragmatic use approximates a
one-to-one form-meaning mapping. As a consequence, a stronger propensity to maximize the
posterior increases their proportion in the population.

However, in contrast to a pressure only for expressivity with high $\lambda$, learnability
alone does not succeed in selecting for a single prevalent type. Indeed, all six target types
tend to coexist at roughly equal proportion. Each is passed on to next generation with the same
faithfulness and, differently from a pressure for communicative success, they do not stand in
competition with each other. As implicit in Figure \ref{fig:only-M}, in $1000$ independent
populations all incumbents were pragmatic $\Llack$ users provided sufficiently high $\lambda$,
with each reaching approximately the same proportion of users in the population. As with
expressivity only, low values of $\lambda$ make the differences in observable behavior across
types less pronounced and therefore reflect the learners' inductive bias more faithfully,
favoring functionally deficient but a priori preferred types such as those that use $\Lall$. A
pressure for learnability alone may consequently lead to a spread of communicatively suboptimal
types that are easier to learn. In the extreme, when $l = 1$ and $\lambda = 1$ all of $1000$
independent populations had users of $\Lall$ as incumbents.

\subsubsection{Combining pressures of expressivity and learnability}

Pressures for communicative success or learnability are not sufficient on their own to have a
single target type dominate the population. When pressured toward expressivity, the slight
communicative advantage of $\Lbound$ users leads to the its prevalence. When pressured toward
learnability, pragmatic $\Llack$ is promoted over functionally similar but semantically more
complex alternatives such as $\Lbound$. However, on its own learnability does not foment the
propagation of these types across the population. What is more, it leads to heterogeneous
populations that foster an equal share of types of a kind, such as different pragmatic $\Llack$
types.

@@

To recapitulate, other than the involvement of pressure on both expressivity and learnability,
the resulting proportion of pragmatic $\Llack$ speakers primarily hinges on three
factors. First, the degree to which agents try to maximize communicative success. This plays a
role both in expressivity as well as in producing data that allows learners to discriminate
this type from others. Second, the inductive bias, which leads learners to prefer simpler over
more complex semantic representations in acquisition. Lastly, the posterior parameter, as it
magnifies the effects of the learning bias in tandem with replication.

\subsection{Summary}

Under the assumption of a learning bias for simpler lexical representations, our results
suggest that a lack of semantic upper-bounds coupled with pragmatic reasoning can overcome
communicative pressures and come to dominate a population. This prediction hinges on three
assumptions. First, that language is pressured toward both expressivity and
learnability. Second, that language users favor expressions/interpretations expected to have a
higher likelihood of being understood/meant over ones with a lower likelihood. Lastly, that
learners prefer simpler over more complex lexical representations and exhibit a tendency
towards the acquisition of the type that best explains the learning data.

\section{General discussion}\label{sec:discussion}

The approach introduced here combines game theoretic models of functional pressure towards
efficient communication \citep{nowak+krakauer:1999}, effects of transmission perturbations on
(iterated) language learning \citep{griffiths+kalish:2007}, probabilistic speaker and listener
types of varied degrees of pragmatic sophistication \citep{frank+goodman:2012,
  franke+jaeger:2014} as well as reasoning about unobservable lexical representations
\citep{bergen+etal:2012,bergen+etal:2016}. This model allows a structured conceptual
investigation of the co-evolution of conventional meaning and pragmatic language use. Main
innovations of the model are (i) its modular separation of expressivity and learnability,
allowing for their isolated and combined analysis, (ii) the characterization of language
learning as a joint inference over types of pragmatic behavior and lexical meaning, and (iii)
the possibility to trace the co-evolution of conventional semantics and pragmatic use.
